\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-15 coq-example.tex
Save. Working on example section

After considering the pros and cons, we decide to implement a Coq Plugin in which we can add new Vernacular command and translate each new command into a bunch of Coq commands (the surface Vernacular command) on the fly, instead of modifying the code-base of Coq.

Despite of the possible difficulty of the future maintenance, this approach has various advantages: 1. it is the easiest and the most accessible way to prototype as it relieves us the necessity of familarity of Coq base, especially for the implementation of the module and functor; 2. we have a clear definition of trusted-base -- the whole Coq; 3. it is easy to debug -- we just need to check the translated commands; 4. it can be well-incorporated with the existent tools like VSCoq; 5. it is more accessible for the the interested audience who can then easily adapt our plugin and give a try -- otherwise they have to download and re-build the whole customized Coq; 6. it is more stable because the surface syntax of Coq should be stable across different versions; 7. still, this plugin can capture the key ingredients of implementing family polymorphism inside Coq and act as a reference for guiding an appropriate implementation of Family Polymorphism inside all sorts of proof assistants.

\textbf{Family compiled into module.} This is the main idea of compiling family components into Coq's primitives: we compile family into coq module, family type into module type and context of a given term into parameters of the module. To achieve incremental type checking, we notice the nature of late binding: for each defining field $ .. \cL\sigma \vdash t : T $, we can compile $t$ using ``universal-quantifier-wrapped term'' : $.. \vdash \lambda t : \forall \cL \sigma. T$. However, instead of using Coq's universal quantifier, we use module and module parameters to achieve this wrapping -- each family type in the context of the judgement will be one module parameter with the compiled module type, and thus each field of the family will compile into a parametrized module (functor). Doing so we can also get rapid feedback from type checking when the users are defining each field interactively with Coq.
% Insert one pseudo-code example for explanation of the mechanism
\begin{figure}
  \begin{minipage}[t]{0.25\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family A.
Field a : nat 
  := 1.
Field b 
  : self_A.a = self_A.a
  := eq_refl.
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.4\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module a_4 (self_A: EmptySig).
Definition a : nat := 1. End a_4.
Module Type a_5 (self_A: EmptySig).
Parameter (a : nat). End a_5.
Module Type a_6 := a_5 EmptyMod.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Module b_7 (self_A: a_6).
Definition b 
  : self_A.a = self_A.a 
  := eq_refl. End b_7.

Module A. (* Final Aggregation *)
Include a_4. Include b_7. End A.
\end{minted}
  \end{minipage}
  \caption{Example Code and Plugin Translation}\label{fig:plugin-example1}
\end{figure}

% Concrete details of compilation of a family term data structure
Taking Figure \ref*{fig:plugin-example1} as an example: when the user interactively input the line "Field a : nat := ..", our plugin will translate this statement into the "Module a_9", "Module Type a_10, a_11" three components. {The "EmptySig" is just empty module type.} Then, these modules and module types will be generated and type-checked by Coq immediately. The computed "Module a_10" is then part of the context of "Field b" (as the module type of "self__A") for "b" to refer to. The internal representation of family "A" is simply the list of these ``compiled'' modules and module types. We can simply achieve incremental checking and overriding -- for example, we want to override "Field a", we simply replace that compiled module with the new module inside the list.

Now that we know a family data structure is actually a list of compiled modules, we may want to aggregate them into one complete compiled module. We achieve that by repeatedly using Coq's "Include" command as the example illustrated. Note that, in our example, we include two functors without arguments -- this is due to one of the feature of "Include": non-instantiated parameters will be automatically instantiated by the appropriate fields in the surrounding defining module. The resulting module is just like mundane Coq's module and thus enjoy all the computational mechanism from Coq for free.

% explaining inheritance
Besides standalone family, we also need to implement family inheritance. To simplify the implementation, we can consider all families as inheritance and the standalone families are extending empty family. Thus we only need to deal with inheritance during interactive proving.

An inheritance data will be annotated with the type for the input family and output family, and generally there are three kinds of inheritance -- extend, override and inherit. Extending corresponds to adding new fields, overriding corresponds to modify existent fields and inheriting corresponds to directly taking fields from the parent. Thus it is possible to consider inheritance as solely a data rather than a functor, and this actually provides us the ability to \text{mix-in} inheritance data.

\textbf{Reasoning Requires Non-Overridable.}
Unlike general-purposed programming, Coq should be able to  \textit{reason} about each program. However, sometimes reasoning requires a specific program and overriding on them is unacceptable. For example, we may define a field $"add" : \mathbb{Z} \to \mathbb{Z} \to \mathbb{Z}$ as one field and intends to proving its commutativity in a following field $"comm"$. But to prove so, we need to know more than just a type interface of $"add"$ but also its \textbf{concrete definition}. Otherwise, in the future, once $"add"$ is overridden into a subtraction function then $"comm"$ will have problem to be  inherited. A more concrete example is in Figure \ref*{fig:plugin-example1}, consider we have "Field b : self_A.a = 1 := .." instead, then our compilation and Coq's type check will both fail: in the context of "b" (module type "a_6") we only know "a" is of type nat but nothing more.  To fix that, we need more information on "a" than its type -- we need to expose its concrete definition "a = 1" to appear in the context of "b".

To achieve such exposure of definition, in meta-theory we only need to expose the information as propositional equality as part of the signature, and in our Coq plugin, we can achieve that even easier by exposing the whole definition directly into module type. Doing so we can have a better judgemental equality and we can avoid overriding because subtraction doesn't have the same definition as $add$. We require the users to decorate field as `Final' so that the plugin will proceed with this special treatment.

\textbf{Special Fields: Overridable and Sealed Family as Interface.}
But there are cases reasoning and overriding are both required, especially when we need to use Coq to do computation. For example, we may want a field to be an arbitrary monotonic function. Again, we cannot split into two fields \mintinline{Coq}{(f : nat → nat, p : monotonic f)} because to provide "p" we need to have concrete definition of "f" in the family type, which makes "f" non-overridable at all. One option is to use sigma type \mintinline{Coq}{{f | monotonic f}} but that is cumbersome when we have a bundle of stuff with multiple refinements.

To resolve this, we provide family sealing mechanism -- basically we allow the users to specify the family/module type, and thus we can have a ``local'' transparent, non-overridable field located in a overridable family. Using the above example, we simply make \mintinline{Coq}{(f : nat → nat, p : monotonic f)} together inside an overridable family, but "f" to be non-overridable and transparent on its definition, and thus we can prove "p". Once we want to swap the implementation, we just override the whole ``sealed'' family.

This ``sealed'' family actually simulate interface to a certain level -- we can assign each field of the sealed family with \mintinline{Coq}{Axiom non_implement : forall {T : Type}, T.}  , and later instantiate them with concrete implementation by overriding this family.
% Basically, {a : int -> int; property : a is monotonic}, in this case
%   we may want to override "a" using different computation, but we still wnat
%   this refinement/property

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We don't actually support extensible inductive type in Coq (because that require non-trivial research effort on theory side -- i.e. {What is the semantic of extensible inductive type?} -- and engineering), but we simulate extensible inductive type using overriding -- when extending an inductive type $A$ with new constructors $c$, our plugin will generate a new inductive definition $A'$ with this new $c$ on the fly and feed to Coq.  But still, we must be careful on the exposing data in the family type (module type) for the following field.

% Insert one pseudo-code example for explanation of the mechanism
% Use the natural number example
\begin{figure}
  \begin{minipage}[t]{0.3\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
Family B.
  FInductive b : Set 
    := tt : b | ff : b.
  Family neg_handler. 
    Field tt : self_B.b 
      := self_B.ff.
    Field ff : self_B.b 
      := self_B.tt.
  EndFamily.
FRecursor neg 
  about self_B.b 
  motive (fun _ => self_B.b)
  using self_B.neg_handler
  by _rec.
EndFamily.

Family B2 extends B.
  Extend FInductive b : Set 
    := uu : b.
  Extend Family neg_handler. 
    Field uu : self_B2.b 
      := self_B2.uu.
  EndFamily. 
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.65\linewidth}
\begin{minted}[fontsize=\footnotesize]{Coq}
  (* Abstraction for Inductive Type *)
Module Type b_3 (self_B: EmptySig_4).
Parameter (b : Set). Parameter (tt ff : b). End b_3.      
Module Type b_6 := b_3 EmptyMod.
  (* Recursor to Set for b *)
Module b_rec_12 (self_B: b_6).
Definition __recursor_type_b_rec :=
  forall P : self_B.b -> Set,
  P self_B.tt -> P self_B.ff 
  -> forall __i : self_B.b, P __i. End b_rec_12. 
  (* Field B.neg_handler.ff *)
Module ff_24 (self_B: b_6) 
             (self__neg_handler: tt_23 self_B).
Definition ff : self_B.b := self_B.tt. End ff_24.
  (* Intermediate Module solely for type checking *)
Module v_33_34 (self_B: neg_handler_26).
Include b_rec_12 self_B.
Parameter (recursor_for_type_checking : __recursor_type_b_rec).
Definition term_for_type_checking :=
  recursor_for_type_checking (fun _ : self_B.b => self_B.b)
    self_B.neg_handler.tt
    self_B.neg_handler.ff.  
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Example Code for Inductive Type}\label{fig:plugin-example2}
\end{figure}

Let's look at \cref{fig:plugin-example2}, the example of boolean and negation. When we define inductive type "b", the following "Field neg_handler.ff" can only know "b" is a type (of "Set") and there are at least two constructors for it, just like what we declare in ``the abstracted interface'' module type "b_3". We cannot export the eliminator of "b" inside "b_3" because doing so "neg_handler.ff" (and "ff_24") will not be able to be inherited to the context where "b" is extended with a third constructor, since the existence of recursor alone will assert that there are only two constructors in "b", not ``extensible'' at all!

But we do have to provide a way to construct a recursor, and execute type checking (or exhaustiveness checking). Thus we need to extract a elimination principle for future usage. Here we show one exemplar extraction of the recursor in \mintinline{Coq}{Set} as module "b_rec_12".

To construct a recursor, we make for each constructor a handler and aggregate them inside a family. The whole family "neg_handler" can be considered as a large matching case. Doing so, family inheritance can function as matching-case reuse. Then we use "FRecursor" to specify the motive and inductive type. Here "FRecursor" will feed Coq
a module "v_33_34" for doing type-checking -- what it does is basically assert the existence of the recursor and apply to see if handlers can pass the type checking by Coq.

Since the recursor is constructed knowing a concrete inductive type, thus not inheritable, for all the children, we have to reconstruct and override with a new recursor (acting like an extra exhaustiveness checking). This is affordable since each recursor handler doesn't need to be rechecked -- look at "B2.neg_handler", there is only one new handler for "uu" and the handlers from the parent "B.neg_handler.tt,B.neg_handler.ff" are already type-checked and inherited(thanks to the fact that they only rely on ``the abstracted interface'' module "b_3"). Of course, our plugin will ``inherits'' the recursor "neg" but what the plugin does is doing exhaustiveness checking, again.

% Explain why the module type need extra care

% explain how recursor is constructed
% explain we have the incremental-checking for any recursor

% explain ftheorem wrapping this complicated recursor
We also provide a command "FTheorem" when the users want to use tactic programming instead. This tool can avoid most boiler-plate code (e.g. specify the handler family) when writing a recursor like above, invoke proof interaction mode, allow tactic usage, and is also open to extension like "FRecursor".

To summarize, the key insight here is that, to make each recursor handler inheritable, we need to seal an abstraction around the inductive type (e.g. module "b_3"); but to construct a recursor (and carry out the exhaustiveness checking), we need to break this abstraction and see the concrete definition of the inductive type (e.g. module "b_rec_12"). Our meta-theory need to handler these two seemingly contradicting ideas simultaneously.

\textbf{Propositional Partial Recursor.}
Apparently, the exposed module "b_3" in \cref*{fig:plugin-example2} is not expressive enough. For example, we cannot prove the constructors are injective, i.e. \mintinline{Coq}{self_B.tt ≠ self_B.ff}. In the original setting, to prove injectiveness of constructor we only require a recursor and its computational rules. However, as we pointed out earlier, the recursor of an inductive type cannot be exposed into the context for the sake of extensibility. We may have to alter the formulation of the recursor.

The simplest idea is to allow partiality -- i.e. returning \mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}. We call this \textit{propositional partial recursor}. For example, the propositional partial recursor for \mintinline{Coq}{B.b} is 
\begin{minted}{Coq}
b_prec : forall P : b -> Set, P tt -> P ff 
    -> forall x : b, option (P x). 
\end{minted}
with appropriate computational axioms. The partiality doesn't break extensibility (i.e. all the future extension of "b" can support this "b_prec").

What's more, partial recursor is theoretically enough for the "injection" and "discriminate" tactics. The rough idea to achieve this is to use "b_prec" to reflect our special inductive type into vanilla inductive type, and use the orignal "injection" and "discrimination" respectively. For example, we want to prove \mintinline{Coq}{self_B.tt ≠ self_B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : self_B.b → option bool 
                   := b_prec (fun _ => bool) true false.
Definition tt_neq_ff : self_B.tt = self_B.ff → False.
(* Because self_B.tt = self_B.ff → reflect self_B.tt = reflect self_B.ff 
                → true = false    (by computational axiom)
                → False           (by discriminate of bool) *)
\end{minted}

In fact, the above "reflect" function implies that propositional partial recursor is good enough to some extent, at least for non-indexed inductive type, because we can ``embed'' vanilla inductive type into types supporting partial recursor. Because "reflect" is actually a left inverse of an injection "bool → self_B.b". A bit more formally, and restrict our focus to non-indexed inductve types,

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e. 
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} that has ...
    \begin{itemize}
      \item following $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor 
      \begin{minted}{Coq}
prec : forall (R : Set), ..,  
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> R) -> R)), .., D -> R
      \end{minted}
      \item all the following $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    \end{itemize}
  \end{itemize}
  We will have an \textbf{embedding} from "C" to "D", more concretely, we can have \mintinline{Coq}{inj : C -> D} and

  \mintinline{Coq}{linv : D -> option C} defined using eliminator and "prec", (both act similar to identity) such that
  \begin{minted}{Coq}
    forall c : C, linv (inj c) = some c
  \end{minted}
\end{theorem}

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can actually be considered as a constructor, because they can be reflected to real constructors of "C" using "linv". Thus, \cref{thm:prec-complete} implies that (1) every future extension of the inductive type can support this partial recursion (trivially); (2) every type support this partial recursor with its computational axiom at least support these constructors because of the embedding. In other words, \textit{a type (at least) supports these constructors if and only if this type supports the corresponding partial recursor (with the computational axiom)}.  Thus we can argue that the partial recursor gives a sound and complete characterization of all the extension of a given (non-indexed) inductive type.

% Mention still needing total recursor, for exhaustiveness checking

\textbf{Hack: Global Reasoning.}
% explain the places using ``Closing Fact''
% 1. No need to export partial recursor
% 2. Computational rule for the total recursor
However, to verify and prove the recursor and partial recursor (and their computational rules), our plugin has to compile the extensible family into a non-extensible Coq Module and prove the recursors by instantiating the content of Coq module with concrete inductive type.

This style of ``Global Reasoning'' is that we can only prove the corresponding theorem when we know the ``full pictures'' of the inductive type. We provide "Closing Fact" as a command to carry out this global reasoning -- we can attach "Closing Fact" with either (proof) terms or Coq's proof script, then only when compilation to module happens, the term/proof script will put into type-check to see if a "Closing Fact" statement can be proved. Note that, "Closing Fact" cannot immeidately check the correctness of the term/proof script because this verification only happens when we finished defining a family, and thus not easy to handle.

This "Closing Fact" can ease us from the necessity to let plugin generate proof for recursors and partial recursors, and thus making plugin development easier. For example, the computation rule for a concrete recursor like "B.neg" and partial recursors in \cref*{fig:plugin-example2} can be directly asserted by using "Closing Fact" and verified later during compilation. This ``Global Reasoning'' will be again useful in the later examples.

\section{Example}
We provide two examples here for illustration and evaluation.

>>>>

We provide two examples here for illustration and evaluation.

### 2022-06-15 coqimpl.tex
Save. Working on example section

This style of ``Global Reasoning'' is that we can only prove the corresponding theorem when we know the ``full pictures'' of the inductive type. We provide "Closing Fact" as a command to carry out this global reasoning -- we can attach "Closing Fact" with either (proof) terms or Coq's proof script, then only when compilation to module happens, the term/proof script will put into type-check to see if a "Closing Fact" statement can be proved. Note that, "Closing Fact" cannot immeidately check the correctness of the term/proof script because this verification only happens when we finished defining a family, and thus not easy to handle.

@@

\section{Example}
We provide two examples here for illustration and evaluation.

\subsection{Type Safety for STLC}
We start with the very basic type safety proof of STLC, largely adapted from Software Foundation, which is also one of the primal motivating example of this project.

Our emphasis here is that, we singled out some programming language features from STLC and prove the type safety for each feature separately, following the style of the examples in MTC\citep{delaware2013,forsta2020}. And we use a \textit{not-yet-completely-defined} \textbf{mix-in} feature to mix the semantic and properties of two programming language feature -- product and boolean -- with vanilla STLC. This way, we can say precisely, \textit{a programming language feature itself is a piece of data/inheritance judgement/inherited family}. Compared to MTC, our example uses small-step operational semantic by exploiting the extensible inductive type, and most of the proof is directly adapted from the one in Software Foundation, resulting a more accessible proof. 
% In related work, Coq/Metatheory a la carte/Tion embedding can be emphasized as a more "semantic approach" because they encode the meaning using a special design pattern (for example, open recursive inductive type for extension), compared to our more syntactic approach. Their advantage is the transferability of this technique accross different proof assistants, and their disadvantage is that their approach are less accessible and unfriendly to amateur Coq users -- which can be reflected from the distance of their approach and text-book Software Foundation proof.

% We have problem on inversion lemma. Check if it is the same problem as MTC
We use `Closing Fact' to state and prove the inversion lemma instead of using the extensible proving mechanism `FTheorem'. The reason is that 1. it introduce much less boilerplate code because the proof for these inversion lemma should be just simple case analysis and we should rely on Coq to auto-generate them; 2. it shouldn't bother us in the future because any extension on the syntax should still satisfy these inversion lemmas; 3. most importantly, we believe this inversion ``lemma'' should be part of the definition of the syntax instead of considering the syntax as a mere concrete inductive type. We should postulate this inversion ``lemma'' like \textit{a constraint} and post-hoc-ly verify that our inductive definition did satisfy the constraint, which is exactly what we expect from `Closing Fact'.

\subsection{Abstract Interpreters for Imp}
The second example is adapted and modified from the Familia\citep{zhang2017familia}
% citation needed here
-- contrary to our first example, we use big-step interpreter and fuel to indicate the operational semantic on an imperative language with side-effect, and we specify the abstract interpretation and prove its soundness, with some of postulation on both computation and property. Then we extend the language feature and we instantiate the postulation on computation for both concrete interpreters and abstract interpreters. Thanks to the compilation, we can directly run the resulting abstract interpreters.

>>>>

This style of ``Global Reasoning'' is that we can only prove the corresponding theorem when we know the ``full pictures'', including the inductive type. We provide "Closing Fact" as a command to carry out this global reasoning -- we can attach "Closing Fact" with either (proof) terms or Coq's proof script, then only when compilation to module happens, the term/proof script will put into type-check to see if a "Closing Fact" statement can be proved. Note that, "Closing Fact" cannot immeidately check the correctness of the term/proof script because this verification only happens when we finished defining a family, and thus not easy to handle.

@@
